---
title: "Matrix Algebra in R Tutorial"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Creating Matrices in R 

A matrix is a two dimensional data structure in R programming. You can create matrices easy in R with the *matrix()* function. 

Please refer to the following video to learn all about creating matrices in R:
![](https://www.youtube.com/watch?v=O7KL17QZNqg)

## Matrix Addition + Subtraction 

```{r prepare-as}
A <- matrix(c(6,2,2,6), nrow =2, ncol=2, byrow = T)
B <- matrix(c(1,3,5,7), nrow =2, ncol=2, byrow = T)
C <- matrix(c(8,4,9,4), nrow =2, ncol=2, byrow = T)
```

*R allows you to perform vectorized operations, which makes matrix operations easy to perform. Three 2x2 matrices **A**, **B**, and **C** have been preloaded for you. The contents of  **A**, **B**, and **C** are shown below for you.*

$$
\mathbf{A} = \left(\begin{array}{cc} 
6 & 2\\
2 & 6
\end{array}\right)
\mathbf{B} = \left(\begin{array}{cc} 
1 & 3\\
5 & 7
\end{array}\right)
\mathbf{C} = \left(\begin{array}{cc} 
8 & 4\\
9 & 4
\end{array}\right)
$$

### Exercise 1: Matrix Addition

Please write R code to perform the following matrix addition: **A** + **B**

```{r as1, exercise = TRUE, exercise.setup = "prepare-as"}

```

```{r as1-solution}
A+B
```

### Exercise 2: Matrix Addition 

Please write R code to perform the following matrix addition: (**A** + **C**) + **B**

```{r as2, exercise = TRUE, exercise.setup = "prepare-as"}

```

```{r as2-solution}
(A+C)+B
```

### Exercise 3: Matrix Subtraction

Please write R code to perfom the following matrix subtraction: **A** - **C**

```{r as3, exercise = TRUE, exercise.setup = "prepare-as"}

```

```{r as3-solution}
A-C
```

### Exercise 4: Matrix Addition and Subtraction

Please write R code to perform the following matrix operation: (**A** - **B**) + **C** 

```{r as4, exercise = TRUE, exercise.setup = "prepare-as"}

```

```{r as4-solution}
(A-B)+C
```

## Matrix Multplication: Part One 

In order to be able to multiply matrices, the number of columns in the first matrix must equal the number of rows in the second matrix. Mathematically speaking, matrices are denoted with *m* rows and *n* columns (m x n). Therefore, the *n* of the first matrix must equal the *m* of the second matrix. 

### Matrix Multiplication Quiz 

```{r mm, echo=FALSE}
quiz(
  question("How many rows does a 4x3 matrix have?",
    answer("4", correct = TRUE),
    answer("3", message = "This is the number of columns")
  ),
  question("How many columns does a 7x8 matrix have?",
    answer("7", message = "This is the number of rows"),
    answer("8", correct = TRUE)
  ),
  question("Can we perform matrix multiplication on two 2x2 matrices?",
    answer("Yes", correct = TRUE),
    answer("No")
  ),
  question("Can we perform matrix multiplication on a 4x3 matrix and a 3x2 matrix?",
    answer("Yes", correct = TRUE),
    answer("No", message = "The number of columns in the first matrix equals the number of rows in
           the second matrix")
  ), 
  question("Can we perform marix multiplication on a 5x6 matrix and a 4x2 matrix?",
    answer("Yes", message = "The number of columns in the first matrix does not equal the number of
           rows in the second matrix"),
    answer("No", correct = TRUE)
)
)
```

## Matrix Multiplication: Part Two

Matrix multiplication is difficult to visualize. To accomplish it, you must multiply the elements of each row of the first matrix by the elements of each column in the second matrix.To visualize this, lets say we have two matrices **A** and **B** from the first exercise

$$
\mathbf{A} = \left(\begin{array}{cc} 
6 & 2\\
2 & 6
\end{array}\right)
\mathbf{B} = \left(\begin{array}{cc} 
1 & 3\\
5 & 7
\end{array}\right)
$$

The result of performing matrix multiplication **A****B** can be visualized in the following way 

$$
\left(\begin{array}{cc}
6*1 + 2*5 & 6*3 + 2*7\\
2*1 + 6*5 & 2*3 + 6*7
\end{array}\right) = 
\left(\begin{array}{cc}
16 & 32\\
32 & 48
\end{array}\right)
$$
Matrices **A** and **B** are already loaded in your working environment. Please run the following code chunk, you can see that the result does not match the expected answer. You cannot use the traditional multiplication symbol in R to perform matrix multiplication. Instead you must use this syntax A %*% B. Now please alter the syntax and confirm that you obtain the right answer. 

### Exercise 5: Matrix Multiplication

```{r mm2, exercise=TRUE, exercise.setup = "prepare-as"}
A*B
```

### Excerise 6: Matrix Multiplication

Now please perform multiplication of **B** and **C** 

```{r mm3, exercise=TRUE, exercise.setup = "prepare-as"}

```

```{r mm3-solution} 
B %*% C
```