---
title: "Matrix Algebra in R Tutorial"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Matrix Addition + Subtraction 

### Matrix Addition 

```{r prepare-as}
A <- matrix(c(3,1,1,4), nrow =2, ncol=2, byrow = T)
B <- matrix(c(7,2,3,5), nrow =2, ncol=2, byrow = T)
C <- matrix(c(3,6,2,4), nrow =2, ncol=2, byrow = T)
```

*R allows you to perform vectorized operations, which makes matrix operations easy to perform. Three 2x2 matrices **A**, **B**, and **C** have been preloaded for you. The contents of matrix **A**, **B**, and **C** are shown below for you.*

$\mathbf{A} = \left(\begin{array}{cc} 
3 & 1\\
1 & 4
\end{array}\right)$

$\mathbf{B} = \left(\begin{array}{cc} 
7 & 2\\
3 & 5
\end{array}\right)$

$\mathbf{C} = \left(\begin{array}{cc} 
3 & 6\\
2 & 4
\end{array}\right)$

Please write R code to perform the following matrix addition: **A** + **B**

```{r as1, exercise = TRUE, exercise.setup = "prepare-as"}

```

```{r as1-solution}
A+B
```

Please write R code to perform the following matrix addition 

```{r as2, exercise = TRUE, exercise.setup = "prepare-as"}

```

### Matrix Subtraction

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```






## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

