---
title: "Principal Component Analysis in R Tutorial"
output: 
  learnr::tutorial:
  progressive: true
  allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## PCA in R

In this tutorial, we will learn how to use the `princomp` function from the `stats` package to perform a principal component analysis (PCA) on an example dataset in R. First, we will examine and prepare the dataset, and then we will analyze it using the `princomp` function.

## Examining the Original Dataset

```{r prepare-data, include = TRUE, echo = TRUE}
seeds_full <- backpack::seeds
seeds <- seeds_full[ , -8]
seeds_standard <- scale(x = seeds, center = TRUE, scale = TRUE)
seeds_PCA <- princomp(seeds_standard, cor = TRUE)
```

We will work with a datset called `seeds`, which contains data on the physical properties of seeds from three species of wheat plants.


### Exercise: Dataset Structure

Complete the R code below to examine the structure of the `seeds` dataset:

```{r str_data, exercise=TRUE, exercise.setup = 'prepare-data'}

```

```{r str_data-solution}
str(seeds)
```

### Exercise: Summary Function

Now call the `summary` function on the dataset to acquire more detailed information on each variable:

```{r summary_data, exercise=TRUE,exercise.setup = 'prepare-data' }
___(seeds)
```

```{r summary_data-solution}
summary(seeds)
```

## Preparing the Dataset

PCA relies on the variance of the variables to create simpler components. However, differing scales and units of measurement across variables may distort our analysis. To overcome this issue, it is generally advisable to center and standardize variables before performing the PCA. 

```{r summary-output, exercise = TRUE , exercise.setup = 'prepare-data'}
print("Mean")
round(apply(seeds, 2, mean), 2)
print("Standard Deviation")
round(apply(seeds, 2, sd), 2)
```


### Exercise: Mean Centering and Standardizing

Centering all variables on their respective mean and then standardizing can be done in one step using the `scale` function from the `base` package included with R. First, variables are centered by subtracting the mean of the variable from each value. Next, the variables are further standardized by dividing each variable by its standard deviation.  

Try standardizing our example dataset using the `scale` function and setting the arguments `center` and `scale` to `TRUE`. Assign this new dataset to the name `seeds_standard`:  

```{r scale_data, exercise=TRUE}
___ <- scale(x = ___, ___, ___)
```

```{r scale_data-solution, exercise.setup = 'prepare-data'}
seeds_standard <- scale(x = seeds, center = TRUE, scale = TRUE)
```


### Exercise: Examining the Standardized Data

Our variables should now all have a mean of zero and units in standard deviations from the mean. Let's run the `summary` function on our standardized data so we can see how it has been transformed:  

```{r summary_standard, exercise = TRUE, exercise.setup = 'prepare-data'}

```

```{r summary_standard-solution}
summary(seeds_standard)
```


## Obtaining the Principal Components

Now that our data is ready, we will run the PCA.

### Exercise: Performing the PCA

Run a principal component analysis using `princomp` on `seeds_standard` and assign it to the name `seeds_PCA`. Make sure to set the argument `cor = TRUE` so that the correlation matrix is used instead of the covariance matrix:  

```{r princomp, exercise=TRUE, exercise.setup = 'prepare-data'}

```

```{r princomp-solution}
seeds_PCA <- princomp(seeds_standard, cor = TRUE)
```

### Exercise: Reading the Summary

Examine the resulting principal components by running `summary` on `seeds_PCA`:  

```{r pca_output, exercise=TRUE}

```

```{r pca_output-solution, exercise.setup = 'prepare-data'}
summary(seeds_PCA)
```

### Exercise: Examining the Loadings

Complete the code below to examine the loadings of the variables on the components from `seeds_PCA`:  

```{r loadings, exercise=TRUE, exercise.setup = 'prepare-data'}
___$loadings
```

```{r loadings-solution}
seeds_PCA$loadings
```


## Plotting Results

Complete the code to create a line plot of the variance of each component from `seeds_PCA`:
 
```{r variance_plot, exercise=TRUE, exercise.setup = 'prepare-data'}
plot(___, type = "l")
```

```{r variance_plot-solution}
plot(seeds_PCA, type = "l")
```



