---
title: "Multivariate Analysis of Variance"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true 
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
# backpack::load_binders("2004-MANOVA")
```

## Introduction to Multivariate Analysis of Variance 

Multivariate Analysis of Variance (MANOVA) is a procedure to analyze group differences for two or more groups. Similar to its univariate counterpart, ANOVA, this multivariate technqiue tests whether or not designated groups appear to differ based on a set of continuous dependent variables considered simultaneously or collectively. 

While ANOVA tests the null hypothesis that all the group means are equal across a dependent variable, MANOVA tests the null hypothesis that all group means are equal, meaning they come from the same population. MANOVA uses a set of metric variables as dependent variables and the objective becomes finding groups of respondents that exhibit differences on the set of dependent variables. Groups are not prespecified, but the researcher uses one or more independent variables to form these groups.

To understand how MANOVA works in a more concrete way, consider the following situation: one's perception of workplace safety and turnover rates. For example, do turnover rates of home-care, school, and hospital-based nurses differ on their perceptions of safety in the workplace. In this case, there are three dependent variables: turnover rates for home-care nurses, turnover rates for school nurses, and turnover rates for hospital-based nurses. We could imagine the independent variable as a sum score from a survey that measures one's perception of workplace safety. In this case, the scores can be categorized into "unsafe", "somewhat unsafe", "neither safe nor unsafe", "somewhat safe", "safe." Therefore, the one independent variable now has five groups. 

## Assumptions of MANOVA 

The helpful assumptions for MANOVA were taken from: https://statistics.laerd.com/stata-tutorials/one-way-manova-using-stata.php. 

This a good source to read for a clear and concrete explanation of MANOVA. 

Important assumptions for MANOVA are:

1. The two or more dependent variables must be continuous 
2. The independent variable must have two or more groups 
3. Each observation can only belong to one group
4. There needs to be more observations in each group than the number of dependent variables
5. No univariate or multivariate outliers
6. Multivariate normality
7. Linear relationship between between each pair of dependent variables for each group of the independent variable 
8. Homogeneity of variance-covariance matrices
9. No Multicollinearity 

### Quiz: Assumptions

```{r quiz1, echo=FALSE}
quiz(
  question("Dependent variables can be categorical ",
    answer("True "),
    answer("False", correct = TRUE)
  ),
  question("Observations can belong to more than one group",
    answer("True"),
    answer("False", correct = TRUE)
  )
)
```

## One-Way MANOVA in R

We will now run through an example of a [one-way] Multivariate Analysis of Variance (MANOVA) in R. The MANOVA function is included in base R, but we will make use of : *dpylr* to manipulate data. The dataset to run the MANOVA is already preloaded for you. The dataset includes 3 variables, two dependent variables and one indepenent variable. The dependent variables are math scores and english scores, and the independent variable is one's anxiety level(low, medium, high). Ultimately, we are trying test whether one's test performance on math and english examinations differed based on test anxiety levels.  

```{r prepare-one}
library(dplyr, quietly = TRUE)
dat <- backpack::anxiety
```

### Step One - Examine Potential Outliers 

The following code allows you to create boxplots for each dependent variable, math + english, at each level of anxiety. Please fill in the *blank* to create boxplots for students with low, med, and high levels of anxiety. 

```{r Outliers, exercise=TRUE, exercise.setup = "prepare-one"}
boxplot(dat$math[dat$anxiety==""])
boxplot(dat$english[dat$anxiety==""])
```

### Step Two - Scatterplots (Linearity)

Outliers were removed from the dataset based on observations in the boxplot. Another important assumption to consider is that there must be a linear relationship between between each pair of the dependent variables for each group of the independent variable.  The following code plots math scores against english scores. You can fill in the blank to show the scatterplot at each level of anxiety. 

```{r prepare-two}
library(dplyr, quietly = TRUE)
dat <- backpack::anxiety
dat = dat %>%
  group_by(anxiety) %>%
  filter(!(abs(math - median(math)) > 2*sd(math)) & !(abs(english-median(english)) > 2*sd(english)))
```

```{r Linearity, exercise = TRUE, exercise.setup="prepare-two"}
plot(dat$math[dat$anxiety==""], dat$english[dat$anxiety==""])
```

### Step Three - Box's M Test (Homogeneity of variance-covariance matrices)

The null hypothesis for this test is that the observed covariance matrices for the 
dependent variables are equal across groups. We want a small enough p-value so that we can reject the null hypothesis and confirm that the covariance matrices for the dependent variables are different across groups. Run the following code to conduct Box's M Test.  

```{r prepare-three}
suppressPackageStartupMessages(library(dplyr, quietly = TRUE))
suppressPackageStartupMessages(library(heplots,quietly = TRUE))

dat <- backpack::anxiety
dat = dat %>%
  group_by(anxiety) %>%
  filter(!(abs(math - median(math)) > 2*sd(math)) & !(abs(english-median(english)) > 2*sd(english)))
```

```{r BoxM, exercise = TRUE, exercise.setup = "prepare-three"}
boxM(dat[,c("math","english")], group = dat$anxiety)
```

### Step Four - Running the MANOVA 

The command to run a MANOVA is already loaded when you startup R as it is part of the stats package. The first line in the code runs the MANOVA. Running the summary.manova command on the fit outputs the results. You can fill in the blank with either Wilks, Pillai, Hotelling-Lawley, or Roy. In this case, we use Wilks. 

```{r MANOVA, exercise = TRUE, exercise.setup = "prepare-three"}
fit <- manova(cbind(math,english)~as.factor(anxiety), data = dat)
summary.manova(fit, test = "Wilks")
```

```{r quiz2, echo=FALSE}
quiz(
  question("Is the p-value for anxiety significant?",
    answer("No"),
    answer("Yes", correct = TRUE)
  )
)
```

## Conclusion 

You can see from the previous exercise that the value in the "Pr(> F)" column is significant, which confirms that the one-way MANOVA result is statistically significant. Therefore, we can confirm that there is a difference in test performance on math and english scores between low, medium and high anxiety students. Overall, a one-way MANOVA only confirms that there is a difference between at least two of the groups. However, it does not tell us exactly which groups are different. Additional, post-hoc testing can be done to find these exact differences, but that is beyond the scope of this lesson. 

